Mathematical Models of Delta-Notch Signalling

### 2025-03-14 paper/paper.tex
Create paper draft with bibliography, generate figures with higher DPI.

\usepackage[left=1.5in, right=1.5in, top=1in, bottom=1in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{graphicx}

@@

@@

\section{Background Information}

Sources: \href{https://www.pnas.org/doi/full/10.1073/pnas.1416287112}{Boareto et al., 2015}, \href{https://www.nature.com/articles/nrm2009}{Bray, 2006}

@@

Binding of the Delta ligand from cell $B$ to the Notch receptor on cell $A$ results in the \emph{cleavage} (removal) of both the ligand and receptor. This interaction triggers the release of NICD (Notch Intracellular Domain) in cell $A$, which \emph{promotes} Notch and \emph{inhibits} Delta.

@@

\section{Ideas and Issues with Collier et al., 1996}

@@

\section{Probabilistic Models of Increasing Complexity}

\textbf{Idea}: Our first goal is to develop a probabilistic model based on a modern understanding of the biochemistry of the delta-notch signalling network. Then, we will use this model to attempt to replicate qualitative results from Collier et al. (1996).

\subsection{A First Model}

@@

\subsection{Computational Implementation}

@@

\subsubsection{Time-Scale Analysis}

The expected time for event $i$ to occur is given by $\mathbb{E}[X_{i}] = 1/\lambda_{i}$. Therefore, the rate at which event $i$ occurs is $\lambda_{i}$, which is equal to the rate at which the event occurs in the original ODE. This means that the trajectories generated from the Gillespie's algorithm implementation of the model can be compared directly to the trajectories generated from the ODE itself, without any need for scaling.

@@

\includegraphics[width=\textwidth]{bifurcation-diagram}

@@

>>>>

\usepackage[left=1.5in, right=1.5in, top=1in, bottom=1in]{geometry}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{hyperref}

\usepackage{graphicx}
\graphicspath{ {./img/} }

\usepackage[backend=biber]{biblatex}
\bibliography{sources.bib}

\title{Mathematical Models of Delta-Notch Signalling}  
\author{Riley Wheadon, Edwin Huras, Ziyi Zhuang}

@@

\maketitle

@@

\section{Abstract}

\section{Introduction}

@@

Binding of the Delta ligand from cell $B$ to the Notch receptor on cell $A$ results in the \emph{cleavage} (removal) of both the ligand and receptor \cite{collier_pattern_1996}. This interaction triggers the release of NICD (Notch Intracellular Domain) in cell $A$, which \emph{promotes} Notch and \emph{inhibits} Delta.

@@

@@

\section{Methods}

\subsection{Parameters and Assumptions}

@@

\subsection{Kolmogorov Forward Equations}

\subsection{Deterministic ODE Model}

\subsection{Stochastic ODE Model}

\subsection{Gillespie Algorithm}

@@

\textbf{Remark}: The expected time for event $i$ to occur is given by $\mathbb{E}[X_{i}] = 1/\lambda_{i}$. Therefore, the rate at which event $i$ occurs is $\lambda_{i}$, which is equal to the rate at which the event occurs in the original ODE. This means that the trajectories generated from the Gillespie's algorithm implementation of the model can be compared directly to the trajectories generated from the ODE itself, without any need for scaling.

@@

\begin{figure}
  \includegraphics[width=\textwidth]{img/bifurcation-diagram.png}
  \caption{Approximate bifurcation diagram of the equilibrium Notch concentration of a single cell under the influence of fixed extracellular Delta from a neighbour.}
\end{figure}

\section{Results}

\subsection{Two-Cell Domain}

\begin{figure}
  \includegraphics[width=\textwidth]{img/two_cell_deterministic.pdf}
  \caption{Results of the two-cell deterministic ODE model.}
\end{figure}

\begin{figure}
  \includegraphics[width=\textwidth]{img/two_cell_stochastic.pdf}
  \caption{Results of the two-cell stochastic ODE model.}
\end{figure}

\begin{figure}
  \includegraphics[width=\textwidth]{img/two_cell_gillespie.pdf}
  \caption{Results of the two-cell Gillespie model.}
\end{figure}

\begin{figure}
  \includegraphics[width=\textwidth]{img/two_cell_comparison.pdf}
  \caption{Comparison of Notch concentrations in deterministic ODE, stochastic ODE, and Gillespie models. All models follow roughly the same trajectory and have similar differentiation times, although the Gillespie model appears to converge slightly faster.}
\end{figure}

\subsection{Linear Domain}

\subsection{Hexagonal Domain}

\section{Discussion}

@@

\nocite{*}
\printbibliography
